---
title: "Simulation of the Neutral Model"
author: "Eric Marcon"
date: "`r Sys.Date()`"
bibliography: references.bib
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
# Installation of packages if necessary
InstallPackages <- function(Packages) {
  InstallPackage <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {
      install.packages(Package, repos="https://cran.rstudio.com/")
    }
  }
  invisible(sapply(Packages, InstallPackage))
}
# Basic packages
InstallPackages(c("bookdown", "formatR", "tidyverse"))
# Specific packages
InstallPackages(c("cli", "entropart", "dbmss", "vegan"))
# knitr options
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, message = FALSE)
# Load the tidyverse
library("tidyverse")
```

# Data

We want to build a species-rich metacommunity and a local, small community that is a sample of the metacommunity, according to the neutral model.

## Species

We will use letters to build the species names, e.g. "AAC".
Combining 3 letters, we'll have $26^3 = 17576$ species napes, i.e. more than the number of species in Amazonia [@TerSteege2013].

To build the names, we use this code:

```{r}
# LETTERS is the vector of capital letters A:Z
# outer() combines the first two args by FUN
outer(LETTERS, LETTERS, FUN = paste0) %>% 
  # add a third letter
  outer(LETTERS, paste0) ->
  species_names
```

## Metacomunity

### Amazonia

The Amazonian metacommunity [@Hubbell2001] follows a log-series distribution [@Fisher1943] with $3.9 \times 10^{11}$ trees and $\alpha = 754$ [@TerSteege2013].

```{r}
mc_size <- 3.9E11
alpha <- 754
# Number of species (Fisher, 1943)
-alpha * log(alpha / (mc_size + alpha)) %>% 
  # integer needed
  as.integer() -> 
  n_species
```

Its estimated number of species is `r format(n_species, scientific = FALSE)`.

We'll draw a random log-series distribution.

```{r}
#' Stochastic abundance of a (single) species in a log-series distribution
#'
#' @param N The size of the metacommunity
#' @param alpha Fisher's alpha, that links the number od species to that of individuals
#'
#' @return The number of individuals of a species
LSabundance <- function(N, alpha) {
  # adapted from Dan Lunn, http://www.stats.ox.ac.uk/~dlunn/BS1_05/BS1_Rcode.pdf
  # Fisher's x is log-series 1-theta
  x <- N / (N + alpha)
  # Draw a random number between 0 and 1
  u <- stats::runif(1)
  # k is the number of individuals to draw
  k <- 1
  # Calculate the probability at k=1
  P <- -x / log(1 - x)
  # Store it in the distribution function
  F <- P
  # Repeat while the cumulated probability is below u
  while (u >= F) {
    # Probability at k+1 obtained from that at k
    P <- P * k * x / (k + 1)
    # Increment k
    k <- k + 1
    # Increment the cumulated probability
    F <- F + P
  }
  return(k)
}
```

Simulating the abundance of the `r format(n_species, scientific = FALSE)` of Amazonia takes a few hours.
The result is saved on disk.
If the file can be found, it is read instead of running the simulation from scratch.

```{r}
if (file.exists("data/amazonia.Rda")) {
  load("data/amazonia.Rda")
} else {
  # Apply LSabundance to each species
  #  replicate(n_species, expr = LSabundance(mc_size, alpha)) does not have a progress bar
  # Loop instead
  library("cli")
  # Prepare a progress bar
  if (interactive()) cli_progress_bar("Log-series", total = n_species)
  # And a vector of abundances
  amz_abundances <- numeric(n_species)
  # Draw each species abundance
  for (s in 1:n_species) {
    amz_abundances[s] <- LSabundance(mc_size, alpha)
    if (interactive()) cli_progress_update()
  }
  save(amz_abundances, file="data/amazonia.Rda")
}

# Set the species names
names(amz_abundances) <- species_names[seq_along(amz_abundances)]

# Declare it is a vector of abundances (package entropart)
library("entropart")
amz_abundances %>% 
  as.AbdVector() ->
  amz_abundances

# Range
summary(amz_abundances)
# Rank-Abundance Curve
autoplot(amz_abundances, Distribution = "lseries")
```

### Barro Colorado Island

We'll use a far smaller dataset according to BCI characteristics [@Hubbell2001]: $\theta=50$ and $m=0.1$.
We first build a metacommunity with a billion trees.
This size is arbitrary: the important parameter is $\theta$.

```{r}
mc_size <- 1E9
# theta is also Fisher's alpha
alpha <- 50
# Number of species (Fisher, 1943)
-alpha * log(alpha / (mc_size + alpha)) %>% 
  # integer needed
  as.integer() -> 
  n_species
```

The metacommunity contains `r n_species` species.
We simulate their abundances.

```{r}
# Apply LSabundance to each species
library("cli")
# Prepare a progress bar
if (interactive()) cli_progress_bar("Log-series", total = n_species)
# And a vector of abundances
mc_abundances <- numeric(n_species)
# Draw each species abundance
for (s in 1:n_species) {
  mc_abundances[s] <- LSabundance(mc_size, alpha)
  if (interactive()) cli_progress_update()
}

# Set the species names
names(mc_abundances) <- species_names[seq_along(mc_abundances)]

# Declare it is a vector of abundances (package entropart)
library("entropart")
mc_abundances %>% 
  as.AbdVector() ->
  mc_abundances

# Range
summary(mc_abundances)
# Rank-Abundance Curve
autoplot(mc_abundances, Distribution = "lseries")
```

## Local community

We draw a local community from the metacommunity.

```{r}
# Choose the size of the community
c_size <- 256
# Draw a random community according to the abundances of the metacommunity
rmultinom(1, size = c_size, prob = mc_abundances) %>% 
  # We need a vector, not a matrix
  as.vector() ->
  c_abundances
# Give the species a name
names(c_abundances) <- names(mc_abundances)
library("entropart")
# Eliminate species with 0 individual
c_abundances[c_abundances > 0] %>% 
  # Declare a vector of abundances
  as.AbdVector() ->
  c_abundances
# Show the RAC
autoplot(c_abundances)
```

# Demographic drift

We set an arbitrary position to the trees of the local community just to visualize it.

```{r}
library("dbmss")
# Build a dataframe for the marks of a point pattern
marks <- data.frame(
  # Species. Each line of the dataframe is an individual
  PointType = rep(names(c_abundances), times = c_abundances), 
  # Size of the points
  PointWeight = rep(1, c_size)
)
# Draw random locations in a square window to obtain a point pattern
c_size %>% 
  runifpoint() ->
  c_ppp
# Add the marks to the point pattern
marks(c_ppp) <- marks
# Make it a wmppp object (package dbmss) to make nice plots
c_wmppp <- as.wmppp(c_ppp)
# Set the factor levels of the point types identical to those of the metacommunity
# so that species of the metacommunity can be used in the local community
c_wmppp$marks$PointType <- factor(
  c_wmppp$marks$PointType, 
  levels = names(mc_abundances)
)
# Map the point pattern
autoplot(c_wmppp)
```

The drift model is as follows: at each step, a tree dies and is replaced by a new tree from the species of the community, with respect to their probabilities.

```{r}
#' One-step evolution of a community without migration
#'
#' @param community A wmppp object (package dbmss) where points are trees.
#' @param show If TRUE, a map of the evolved community is produced.
#'
#' @return A wmppp object with the modified community.
evolve_drift <- function(community, show = FALSE) {
  # Choose a tree
  focus_tree <- sample(seq_len(community$n), size = 1)
  # Species of the survivors
  survivors_sp <- community$marks$PointType[-focus_tree]
  # Draw the species of the new tree
  community$marks$PointType[focus_tree] <- 
    survivors_sp[sample(seq_along(survivors_sp), size = 1)]
  if (show) {
    # Copy the community
    community_to_plot <- community
    # Increase the size of the modified tree
    community_to_plot$marks$PointWeight[focus_tree] <- 2
    # Plot the community (print is mandatory inside a function)
    print(autoplot(community_to_plot))
  }
  # return
  return(community)
}

# Test the function
c_wmppp %>% 
  evolve_drift(show = TRUE)
```

Drift causes the local extinction of all species but one.

```{r}
library("cli")
# Copy the community to modify it
the_community <- c_wmppp
# Number of species
richness <- length(unique(the_community$marks$PointType))
# Prepare a long vector to save the richness along time
richness_time <- numeric(1E6)
# Count the steps
step <- 0
# Prepare a progress bar
if (interactive()) cli_progress_bar("Drifting")
while (richness > 1) {
  step <- step + 1
  # Replace a tree
  the_community <- evolve_drift(the_community)
  # Update the richness
  richness <- length(unique(the_community$marks$PointType))
  # Save it
  richness_time[step] <- richness
  # Show the progress
   if (interactive()) cli_progress_update()
}
# Finalise the progress bar
if (interactive()) cli_progress_update(force = TRUE)
# Save the necessary time to reach a single species
drift_time <- step
# Plot the community
autoplot(the_community)
# Plot the decreasing richness
data.frame(Time = seq_len(step), Richness = richness_time[seq_len(step)]) %>% 
  ggplot() +
    geom_line(aes(x = Time, y = Richness))
```

# Immigration

Immigration is the ability to replace the dead tree in the local community by a tree of a species of the metacommunity.

The parameter `m` is the probability for that to happen.
The evolution function is rewritten.

```{r}
#' One-step evolution of a community with migration
#'
#' @param community A wmppp object (package dbmss) where points are trees.
#' @param metacommunity A named vector of abundances or probability that describes the metacommunity.
#' @param m The migration parameter, that is the probability for a dead tree to be replaced by an immigrant species.
#'
#' @return A wmppp object with the modified community.
evolve_migrate <- function(community, metacommunity, m = 0) {
  # Choose a tree
  focus_tree <- sample(seq_len(community$n), size = 1)
  # Migration or local parent? Draw in a uniform distribution.
  u <- runif(1)
  if (u > m) {
    # Local parent: Species of the survivors
  survivors_sp <- community$marks$PointType[-focus_tree]
  # Draw the species of the new tree
  community$marks$PointType[focus_tree] <- 
    survivors_sp[sample(seq_along(survivors_sp), size = 1)]
  } else {
    # Migration : Draw the species of the new tree in the metacommunity
    community$marks$PointType[focus_tree] <- 
      names(metacommunity)[which(rmultinom(1, size = 1, prob = metacommunity) == 1)]
  }
  # return
  return(community)
}
```

We make the community evolve the same amount of time as needed for complete drift and see what happens.

```{r}
# Choose the migration probability
m <- 0.1

# Copy the community to modify it
the_community <- c_wmppp
# Number of species
richness <- length(unique(the_community$marks$PointType))
# Prepare a long vector to save the richness along time
richness_time <- numeric(1E6)
# Prepare a progress bar
if (interactive()) cli_progress_bar(paste("Migration", m, ":"), total = drift_time)
for (step in seq_len(drift_time)) {
  # Replace a tree
  the_community <- evolve_migrate(the_community, mc_abundances, m)
  # Save the richness
  richness_time[step] <- length(unique(the_community$marks$PointType))
  # Show the progress
  if (interactive()) cli_progress_update()
}
autoplot(the_community)
# Plot richness
data.frame(Time = seq_len(drift_time), Richness = richness_time[seq_len(drift_time)]) %>% 
  ggplot() +
    geom_line(aes(x = Time, y = Richness))
# Plot the rank-abundance curve
the_community %>% 
  # Consider the dataframe of marks
  marks() %>% 
  # Group by species
  group_by(PointType) %>% 
  # Count the numnber of individuals by species
  summarise(Abundance = n()) %>% 
  # Extract the column 
  pull(Abundance) %>% 
  # Make it an abundance vector (package entropart)
  as.AbdVector() %>% 
  # Plot the rank-abundance curve and try to fit a log-normal distribution
  autoplot(Distribution = "lnorm")
```

# Exercices

Re-use the code above to run a few experiments about drift time and the shape of the local community.

## The immigration parameter

From the BCI metacommunity, draw a local 1-ha community.
Choose two immigration rates in turn: 0.1, as observed at BCI, and 0.01.
Run a 50,000 steps and observe the final RAC.
Compare the nuber of species with the actual value.

BCI data is available in package vegan.

```{r}
library("vegan")
data(BCI)
# Number of trees per ha
sum(BCI)/50
# Number of trees in each 1-ha plot
rowSums(BCI)
# Number of species in each 1-ha plot
rowSums(BCI > 0)
# RAC of plot 1
BCI[1, ] %>%
  as.AbdVector() %>% 
  autoplot()
```



## The effect of the size of the local community to drift

Draw small communities with 10 species, in log-normal distributions, ignoring the metacommunity.
Make their size vary: 16, 32, 64, 128, 256 individuals.
Run the drift model and note the necessary time for complete drift.
Compare your results with figure 4.12 (page 106) of @Hubbell2001.

```{r}
# Hint
rCommunity(1, size = 256) %>% 
  autoplot()
```

# References
