---
title: "Simulation of the Neutral Model"
author: "Eric Marcon"
date: "`r Sys.Date()`"
bibliography: references.bib
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r}
#| label: Setup
#| include: false

# Packages to install
packages_required <- c(
  # Tools, do not remove them
  "bookdown", 
  "formatR", 
  "tidyverse", 
  # Specific packages
  "cli", 
  "divent", 
  "vegan"
)

# Already installed?
packages_installed <- vapply(
  packages_required, 
  FUN = requireNamespace, 
  FUN.VALUE = TRUE, 
  quietly = TRUE
) 
# Install missing packages
install.packages(
  names(packages_installed)[!packages_installed],
  repos = "https://cran.rstudio.com/",
  quiet = TRUE
)

# knitr options
knitr::opts_chunk$set(
  echo = TRUE, 
  cache = TRUE, 
  message = FALSE
)

# Random seed
set.seed(973)
```

# Data and tools

We want to build a species-rich metacommunity and a local, small community that is a sample of the metacommunity, according to the neutral model.

## Species

We will use letters to build the species names, e.g. "AAC".
Combining 3 letters, we'll have $26^3 = 17576$ species names, i.e. more than the number of species in Amazonia [@TerSteege2013].

To build the names, we use this code:

```{r}
library("tidyverse")
# LETTERS is the vector of capital letters A:Z
# outer() combines the first two args by FUN
LETTERS %>% 
  outer(LETTERS, FUN = paste0) %>% 
  # add a third letter
  outer(LETTERS, FUN = paste0) ->
  species_names
```


## Barro Colorado Island

We build a dataset according to BCI characteristics [@Hubbell2001]: $\theta=50$ and $m=0.1$.
The metacommunity follows a log-series distribution [@Fisher1943].
We build it with a billion trees.
This size is arbitrary: the important parameter is $\theta$.

```{r}
mc_size <- 1E9
# theta is also Fisher's alpha
alpha <- 50
# Number of species (Fisher, 1943)
(-alpha * log(alpha / (mc_size + alpha))) %>% 
  # integer needed
  as.integer() -> 
  n_species
```

The metacommunity contains `r n_species` species.
We simulate their abundances.

```{r}
library("divent")
rcommunity(
  # Simulate a single community
  1,
  # Number of trees
  size = mc_size,
  # Distribution: log-series
  distribution = "lseries",
  # Parameter
  fisher_alpha = alpha
) %>% 
  # Set the species names 
  rename_with(
    # Function of the selected columns: 
    # return a subset of the species names vector
    function(x) species_names[seq_along(x)],
    # Select columns with species names
    starts_with("sp_")
  ) ->
  mc_abundances

# Rank-Abundance Curve
autoplot(mc_abundances, Distribution = "lseries")
```

## Local community

We draw a local community from the metacommunity.
We give random positions to the trees in a square window to be able to see them later.

```{r}
# Choose the size of the community
c_size <- 256
# Draw a random community according to the abundances of the metacommunity
rspcommunity(
  # Draw 1 community
  1,
  # Number of trees
  size = c_size,
  # Metacommunity abundances, must be a vector
  abd = as.numeric(mc_abundances),
  # Draw a multinomial distribution
  bootstrap = "Marcon2012",
  # Random location
  spatial = "Binomial"
  ) ->
  # The result is a weighted, marked, planar point pattern
  c_wmppp

# Show the RAC
c_wmppp %>% 
  # Make it an abundances object
  as_abundances() %>% 
  autoplot()
```

# Demographic drift

## Map the point pattern

```{r}
autoplot(c_wmppp) +
  labs(colour = "Species", size = "Size")
```

The drift model is as follows: at each step, a tree dies and is replaced by a new tree from the species of the community, with respect to their probabilities.

```{r}
library("dbmss")
#' One-step evolution of a community without migration
#'
#' @param community A wmppp object (package dbmss) where points are trees.
#' @param show If TRUE, a map of the evolved community is produced.
#'
#' @return A wmppp object with the modified community.
evolve_drift <- function(community, show = FALSE) {
  # Choose a tree
  focus_tree <- sample(seq_len(community$n), size = 1)
  # Species of the survivors
  survivors_sp <- marks(community)$PointType[-focus_tree]
  # Draw the species of the new tree
  marks(community)$PointType[focus_tree] <- 
    survivors_sp[sample(seq_along(survivors_sp), size = 1)]
  if (show) {
    # Copy the community
    community_to_plot <- community
    # Increase the size of the modified tree
    marks(community_to_plot)$PointWeight[focus_tree] <- 2
    # Plot the community (print is mandatory inside a function)
    print(autoplot(community_to_plot))
  }
  # return
  return(community)
}

# Test the function
c_wmppp %>% 
  evolve_drift(show = TRUE) +
  theme(legend.position = "none")
```

Drift causes the local extinction of all species but one.

```{r}
library("cli")
# Copy the community to modify it
the_community <- c_wmppp
# Number of species
richness <- length(unique(marks(the_community)$PointType))
# Prepare a long vector to save the richness along time
richness_time <- numeric(1E6)
# Count the steps
step <- 0
# Prepare a progress bar
if (interactive()) cli_progress_bar("Drifting")
while (richness > 1) {
  step <- step + 1
  # Replace a tree
  the_community <- evolve_drift(the_community)
  # Update the richness
  richness <- length(unique(marks(the_community)$PointType))
  # Save it
  richness_time[step] <- richness
  # Show the progress
   if (interactive()) cli_progress_update()
}
# Finalise the progress bar
if (interactive()) cli_progress_update(force = TRUE)
# Save the necessary time to reach a single species
drift_time <- step
# Plot the community
autoplot(the_community)
# Plot the decreasing richness
data.frame(Time = seq_len(step), Richness = richness_time[seq_len(step)]) %>% 
  ggplot() +
  geom_line(aes(x = Time, y = Richness))
```

# Immigration

Immigration is the ability to replace the dead tree in the local community by a tree of a species of the metacommunity.

The parameter `m` is the probability for that to happen.
The evolution function is rewritten.

```{r}
#' One-step evolution of a community with migration
#'
#' @param community A wmppp object (package dbmss) where points are trees.
#' @param metacommunity A species_distribution object (package divent) that describes the metacommunity.
#' @param m The migration parameter, that is the probability for a dead tree to be replaced by an immigrant species.
#'
#' @return A wmppp object with the modified community.
evolve_migrate <- function(community, metacommunity, m = 0) {
  # Choose a tree
  focus_tree <- sample(seq_len(community$n), size = 1)
  # Migration or local parent? Draw in a uniform distribution.
  u <- runif(1)
  if (u > m) {
    # Local parent: Species of the survivors
    survivors_sp <- marks(community)$PointType[-focus_tree]
    # Draw the species of the new tree
    marks(community)$PointType[focus_tree] <- 
      survivors_sp[sample(seq_along(survivors_sp), size = 1)]
  } else {
    # Metacommunity species names
    sp_names <- colnames(prob_species(metacommunity))
    # Metacommunity species abundances
    sp_abd <- as.numeric(metacommunity)
    # Draw a species in the metacommunity:
    # a vector of species abd with a single non-zero value
    mc_1_tree <- rmultinom(1, size = 1, prob = sp_abd)
    # Set the new species
    marks(community)$PointType[focus_tree] <- sp_names[mc_1_tree]
  }
  # return
  return(community)
}
```

We make the community evolve the same amount of time as needed for complete drift and see what happens.

```{r}
# Choose the migration probability
m <- 0.1

# Copy the community to modify it
the_community <- c_wmppp
# Prepare a long vector to save the richness along time
richness_time <- numeric(drift_time)
# Prepare a progress bar
if (interactive()) {
  cli_progress_bar(paste("Migration", m, ":"), total = drift_time)
}
for (step in seq_len(drift_time)) {
  # Replace a tree
  the_community <- evolve_migrate(the_community, mc_abundances, m)
  # Save the richness
  richness_time[step] <- length(unique(marks(the_community)$PointType))
  # Show the progress
  if (interactive()) cli_progress_update()
}
autoplot(the_community)
```

```{r}
# Plot richness along time
data.frame(
  Time = seq_len(drift_time), 
  Richness = richness_time
  ) %>% 
  ggplot() +
  geom_line(aes(x = Time, y = Richness))
```

```{r}
# Plot the rank-abundance curve
the_community %>% 
  # Make it an abundance object (package divent)
  as_abundances() %>% 
  # Plot the rank-abundance curve and try to fit a log-normal distribution
  autoplot(distribution = "lnorm", fit_rac = TRUE)
```

The average number of species can be calculated from the data, after eliminating the burn-in steps (i.e. before the pattern is stable).

```{r}
mean(richness_time[10000:drift_time])
```
The equation for the expected number of species is in @Etienne2007 [eq. 4] but it requires knowing the size of the metacommunity.

The minimum richness is `r min(richness_time[10000:drift_time])`.



# Exercises

Re-use the code above to run a few experiments about drift time and the shape of the local community.

## The immigration parameter

From the BCI metacommunity, draw a local 1-ha community.
Choose two immigration rates in turn: 0.1, as observed at BCI, and 0.01.
Run 50,000 steps and observe the final RAC.
Compare the number of species with the actual value.

BCI data is available in package vegan.

```{r}
library("vegan")
data(BCI)
# Number of trees per ha
sum(BCI)/50
# Number of trees in each 1-ha plot
rowSums(BCI)
# Number of species in each 1-ha plot
rowSums(BCI > 0)
# RAC of plot 1
BCI[1, ] %>%
  as_abundances %>% 
  autoplot()
```



## The effect of the size of the local community to drift

Draw small communities with 10 species, in log-normal distributions, ignoring the metacommunity.
Make their size vary: 16, 32, 64, 128, 256 individuals.
Run the drift model and note the necessary time for complete drift.
Compare your results with figure 4.12 (page 106) of @Hubbell2001.

```{r}
# Hint
rcommunity(1, size = 256) %>% 
  autoplot()
```

# References
